Echoes from the scattering of wavepackets on wormholes

### 2019-01-19 paper2.tex
Now I'm done with it...

Amplitudes of the outgoing signals depend on the width of the incident Gaussian wavepacket, this is visible in Fig.~\ref{fig:asympt_two_sigmas} where we plot the asymptotic solutions for two different ingoing wavepackets: one with $\sigma=0.6495r_g$ in the left panel and a second one with $\sigma=5.196r_g$ in the right. After subtracting the outgoing solution for a black hole, the presence of a train of wavelets, colloquially known as echoes, is very clear. For large values of $\sigma$, the signal is not featureless after the transient, as we can observe in the right panel of the same figure. Therefore, subtracting the outgoing pulse (i.e., the case in which there is only one potential wall) obtained from the black hole is a convenient way to clean the signal from back reflections due to the ``tails'' of the potential barrier. In the particular case depicted in the right panel of Fig.~\ref{fig:asympt_two_sigmas}, the amplitudes of the echoes are four orders of magnitude smaller than the transient. In both panels, we plot the variable $\Delta\mathcal{U}_{20}|_{\mathrm{clean}}\equiv\mathcal{U}_{20}|_{\mathrm{original}}-\mathcal{U}^{\mathrm{bh}}_{20}$ in the right upper corner of the figures to represent the echoes and their net amplitude after removing backscattering effects. Notice that, in the upper corner of both figures, the amplitude of the echoes does not decrease exponentially with time due to the successive reflections inside the cavity.

As shown in subsection \ref{subsec:bh}, the curves of reflectivity and transmissivity determine the frequencies going through the barriers of the potential cavity: most of the power of an incident pulse with large $\sigma$ is in the low frequency domain, and therefore it will be reflected. Moreover, the cavity is transparent to high frequency signals, which are dominant in the pulses with small $\sigma$. In either of these extremal scenarios, QNMs cannot be sourced by internal reflections and thus, the amplitude of the echoes is not large in general.\\

@@

Now we proceed to reconstruct the asymptotic spectrum by following the geometrical optics relation in \eqref{eq:reconst} considering the reflectivity and transmissivity operators defined in \eqref{eq:ref_and_trans}. Henceforth, the outcome should be compared with the spectral content of the asymptotic wave solutions of \eqref{eq:wave_scalar_wh}. We calculate the Fourier transforms of both the Gaussian incident pulse $\Phi_0(\omega)=\mathcal{F}[\exp\left((x-x_0)^2/2\sigma^2\right)]$ for $\sigma=0.6495r_g$ and $x_0=60.0r_g$, and the asymptotic solution $\tilde{\mathcal{U}}_{20}^{\mathrm{echo}}\equiv\mathcal{F}[\mathcal{U}_{20}(t,+\infty)]$ including the echoes.\\

@@

We show our results in Fig.~\ref{fig:echo_sigma}, where we notice in the left panel the presence of a well-defined maximal amplitude of the first four echoes. This is consistent with the idea of a range of frequencies/widths that maximize the amplitudes of the echoes, as we shown in the analysis of the reflection and transmission coefficients: the cavity is transparent for small widths of the ingoing Gaussian (large frequencies) and is reflective for wide incident pulses. In the right panel, we observe the growth of the relative amplitude as the widths become smaller. Such a fact only means that the reduction of the transient is faster than the reduction of the echoes as the frequencies grow, as the cavity becomes more transparent the ingoing pulses get transmitted more efficiently. Therefore, this analysis is useful to get a basic understanding of the frequency/width scales in which echoes could be observable.

@@

\section{Scattering of tensor wavepackets}\label{sec:tensor} 
Following the perspective of the even/odd parity decomposition for tensor perturbations of a spherically symmetric spacetime \citep{Regge:1957td, PhysRevD.2.2141, PhysRevD.5.2419}, we implement all the techniques used in section \ref{sec:scalar} and extend our analysis to study the scattering of a test Gaussian wavepacket of tensor perturbations. From now on, we will follow the conventions in \citep{Martel:2005ir}, including the choice of the Regge-Wheeler gauge. 
The dynamics of the wave scattering problem is given by two equations of motion of the form
\left(\tilde{\Box}-\mathcal{V}_{\mathrm{eff}}\right)\Psi_{\ell m}=0,
\label{eq:tensor_eq_mov}
here $\tilde{\Box}\equiv g^{ab}\nabla_a\nabla_b$ is the 2D d'Alembertian operator in the usual $(a,b)\rightarrow(t,r)$ Schwarzschild coordinates. $\mathcal{V}_{\mathrm{eff}}$ corresponds to one of two possible potentials, the Regge-wheeler (odd) potential, $\mathcal{V}_{\mathrm{odd}}$ 
\mathcal{V}_{\mathrm{odd}}(r)= \frac{\ell(\ell+1)}{r^2}-\frac{3r_g}{r^3},
\label{eq:V_odd}
or the Zerilli (even) potential, $\mathcal{V}_{\mathrm{even}}$ 
\mathcal{V}_{\mathrm{even}}(r)=\displaystyle{\frac{1}{\Lambda^2}\bigg[\mu^2\bigg(\frac{\mu+2}{r^2}+\frac{3r_g}{r^3}\bigg)+\frac{9r_g^2}{r^4}\left(\mu+\frac{r_g}{r}\right)\bigg]},
\label{eq:V_even}
where $\mu\equiv(\ell-1)(\ell+2)$ and $\Lambda\equiv\mu+3r_g/r$. All the source terms proportional to the stress energy tensor and its contractions appearing in the right hand side of \eqref{eq:tensor_eq_mov} in \citep{Martel:2005ir} are not considered for the scattering problem. The introduction of tortoise coordinates $(t,x)$ is also very convenient and works in exactly the same way as in \eqref{eq:tortoise} and \eqref{eq:tortoise2}, in these coordinates the waveform of the two equations of motion -- one for the odd parity modes and another for the even -- is given by 
\left[-\partial_t^2+\partial_x^2-V_{\mathrm{eff}}(x)\right]\Psi_{\ell m}(x,t) = 0,
\label{eq:wave_tensor_wh}
where $V_{\mathrm{even}}\equiv[1-r_g/r(|x|)]\mathcal{V}_{\mathrm{even}}$ and $V_{\mathrm{odd}}\equiv[1-r_g/r(|x|)]\mathcal{V}_{\mathrm{odd}}$. Recalling the procedure followed in subsection \ref{subsec:wh}, our setup already includes the effective potentials for the Morris-Thorne wormholes, obtained by reflection of the potential barriers about the ordinate axis. The effective potentials are plotted in the left panel of Fig.~\ref{fig:evenodd_potentials}, which is very similar to the one in the scalar scattering. In the right panel it is possible to notice that the difference between the potentials is only substantial at $(\ell\leq1)$. Both the even and odd solutions of \eqref{eq:tensor_eq_mov} and \eqref{eq:wave_tensor_wh} are already spherical modes used to find the two asymptotic polarizations of the tensor fluctuations propagating in a flat background, such as the term in the diagonal, $h_+$ 
\begin{align}
\displaystyle{h_+=\frac{1}{r(|x|)}\sum_{\ell,m}\bigg\{\Psi_{\ell m}^{\mathrm{even}}\left[\partial^2_{\theta}+\frac{1}{2}\ell(\ell+1)\right]Y_{\ell m}(\theta,\phi)}\nonumber\\
\displaystyle{-\Psi_{\ell m}^{\mathrm{odd}}\frac{i m}{\sin\theta}\left[\partial_{\theta}-\frac{\cos\theta}{\sin\theta}\right]Y_{\ell m}(\theta,\phi)\bigg\}},
\label{eq:hplus}
\end{align} 
and the off-diagonal, $h_{\times}$ 
\begin{align}
\displaystyle{h_{\times}=\frac{1}{r(|x|)}\sum_{\ell,m}\bigg\{\Psi_{\ell m}^{\mathrm{odd}}\left[\partial^2_{\theta}+\frac{1}{2}\ell(\ell+1)\right]Y_{\ell m}(\theta,\phi)}\nonumber\\
\displaystyle{+\Psi_{\ell m}^{\mathrm{even}}\frac{i m}{\sin\theta}\left[\partial_{\theta}-\frac{\cos\theta}{\sin\theta}\right]Y_{\ell m}(\theta,\phi)\bigg\}}.
\label{eq:hx}
\end{align}
From these expressions, it is simple to see that the the monopole ($\ell=0$) and the dipole ($\ell=1$) terms are identically zero. Thus, the first nontrivial contributions come from the quadrupole solutions $\Psi_{20}^{\mathrm{odd}}(x,t)$ and $\Psi_{20}^{\mathrm{even}}(x,t)$, from which the differences in the odd and even potentials are small, and become even smaller for every $\ell>3$, as we can see in right panel of Fig.~\ref{fig:evenodd_potentials}. Additionally, it is reasonable to identify $h_+$ with the even mode and  $h_{\times}$ with the odd in the equatorial plane up to a constant. Hence, we can study any of these two contributions since it is not necessary to duplicate solutions of the mode equations in order to extend our discussions from section \ref{sec:scalar}. Nonetheless, we will explain one of the consequences of the difference between the Regge-Wheeler and Zerilli potentials in Appendix \ref{sec:AppendixPolar}.\\

@@

\section{Conclusions}\label{sec:conclusions}
In this paper, we studied the scattering of a test scalar and tensor wavepacket on a Morris-Thorne wormhole. Using a Gaussian pulse as an incident initial condition, we showed the time dependent scattering solution of the quadrupole scalar and tensor spherical modes both inside and outside the potential cavities in Figs.~\ref{fig:sol_n_flux_wh} and \ref{fig:even_evolve}, which reflect and transmit throughout the cavity. Furthermore, after finding the transmission and reflection coefficients of the cavities in Fig.~\ref{fig:RT_scalar} and in the left panel of Fig.~\ref{fig:RT_reconst}, we used the geometrical optics approximation to reconstruct the shape of the Fourier transformed asymptotic solutions in Fig.~\ref{fig:rec_scalar} and in the right panel of Fig.~\ref{fig:RT_reconst}. We find that the reconstructed shape of the spectrum is accurate, without showing, however, the quasinormal mode peaks.

>>>>

Amplitudes of the outgoing signals depend on the width of the incident Gaussian wavepacket, this is visible in Fig.~\ref{fig:asympt_two_sigmas} where we plot the asymptotic solutions for two different ingoing wavepackets: one with $\sigma=0.6495r_g$ in the left panel and a second one with $\sigma=5.196r_g$ in the right. After subtracting the outgoing solution for a black hole, the presence of a train of wavelets, colloquially known as echoes, is very clear. For large values of $\sigma$, the signal is not featureless after the transient, as we can observe in the right panel of the same figure. Therefore, subtracting the outgoing pulse (i.e., the case in which there is only one potential wall) obtained from the black hole is a convenient way to clean the signal from back reflections due to the ``tails'' of the potential barrier. The necessity of this procedure is more evident in the case depicted in the right panel of Fig.~\ref{fig:asympt_two_sigmas}, where the amplitudes of the echoes are four orders of magnitude smaller than the transient. In both panels, we plot the variable $\Delta\mathcal{U}_{20}|_{\mathrm{clean}}\equiv\mathcal{U}_{20}|_{\mathrm{original}}-\mathcal{U}^{\mathrm{bh}}_{20}$ in the right upper corner of the figures to represent the echoes and their net amplitude after removing backscattering effects. Notice that, in the upper corner of both figures, the amplitude of the echoes does not decrease exponentially with time due to the successive reflections inside the cavity.

As shown in subsection \ref{subsec:bh}, the curves of reflectivity and transmissivity determines which frequencies stay in the cavity: most of the power of an incident pulse with large $\sigma$ is in the low frequency domain, and therefore it will be reflected. The cavity is transparent to high frequency signals, which are dominant in the pulses with small $\sigma$. In either of these extremal scenarios, QNMs cannot be sourced by internal reflections and thus, the amplitude of the echoes is not large in general. Furthermore, the steepness of the transition near the overlap point $R^2(\omega)=T^2(\omega)=0.5$ regulates the abundance of frequencies in the spectrum of outgoing signals.\\

@@

As a next step, we proceed to reconstruct the asymptotic spectrum by following the geometrical optics relation in \eqref{eq:reconst} considering the reflectivity and transmissivity operators defined in \eqref{eq:ref_and_trans}. Henceforth, the outcome should be compared with the spectral content of the asymptotic wave solutions of \eqref{eq:wave_scalar_wh}. We calculate the Fourier transforms of both the Gaussian incident pulse $\Phi_0(\omega)=\mathcal{F}[\exp\left((x-x_0)^2/2\sigma^2\right)]$ for $\sigma=0.6495r_g$ and $x_0=60.0r_g$, and the asymptotic solution $\tilde{\mathcal{U}}_{20}^{\mathrm{echo}}\equiv\mathcal{F}[\mathcal{U}_{20}(t,+\infty)]$ including the echoes.\\

@@

Our results can be found in Fig.~\ref{fig:echo_sigma}, where we notice in the left panel the presence of a well-defined maximal amplitude of the first four echoes. This is consistent with the idea of a range of frequencies/widths that maximize the amplitudes of the echoes, as shown in the analysis of the reflection and transmission coefficients: the cavity is transparent for small widths of the ingoing Gaussian (large frequencies) and is reflective for wide incident pulses. In the right panel, we observe the growth of the relative amplitude as the widths become smaller. Such a fact only means that the reduction of the transient is faster than the reduction of the echoes as the frequencies grow, as the cavity becomes more transparent the ingoing pulses get transmitted more efficiently. Consequently, this analysis is useful to get a basic understanding of the frequency/width scales in which echoes could be observable.

@@

\section{Scattering of tensor wavepackets}\label{sec:tensor} 
Following the perspective of the even/odd parity decomposition for tensor perturbations of a spherically symmetric spacetime \citep{Regge:1957td, PhysRevD.2.2141, PhysRevD.5.2419}, we implement all the techniques used in section \ref{sec:scalar} and extend our analysis to study the scattering of a test Gaussian wavepacket of tensor perturbations. From now on, we will follow the conventions in \citep{Martel:2005ir}, including the choice of the Regge-Wheeler gauge. 
The dynamics of the wave scattering problem is given by two equations of motion of the form
\begin{equation}
\left(\tilde{\Box}-\mathcal{V}_{\mathrm{eff}}\right)\Psi_{\ell m}=0,
\label{eq:tensor_eq_mov}
\end{equation} 
here $\tilde{\Box}\equiv g^{ab}\nabla_a\nabla_b$ is the 2D d'Alembertian operator in the usual $(a,b)\rightarrow(t,r)$ Schwarzschild coordinates. $\mathcal{V}_{\mathrm{eff}}$ corresponds to one of two possible potentials, the Regge-wheeler (odd) potential, $\mathcal{V}_{\mathrm{odd}}$ 
\begin{equation}
\mathcal{V}_{\mathrm{odd}}(r)= \frac{\ell(\ell+1)}{r^2}-\frac{3r_g}{r^3},
\label{eq:V_odd}
\end{equation} 
or the Zerilli (even) potential, $\mathcal{V}_{\mathrm{even}}$ 
\begin{equation}
\mathcal{V}_{\mathrm{even}}(r)=\displaystyle{\frac{1}{\Lambda^2}\bigg[\mu^2\bigg(\frac{\mu+2}{r^2}+\frac{3r_g}{r^3}\bigg)+\frac{9r_g^2}{r^4}\left(\mu+\frac{r_g}{r}\right)\bigg]},
\label{eq:V_even}
\end{equation} 
where $\mu\equiv(\ell-1)(\ell+2)$ and $\Lambda\equiv\mu+3r_g/r$. All the source terms proportional to the stress energy tensor and its contractions appearing in the right hand side of \eqref{eq:tensor_eq_mov} in \citep{Martel:2005ir} are not considered for the scattering problem. The introduction of tortoise coordinates $(t,x)$ is also very convenient and works in exactly the same way as in \eqref{eq:tortoise} and \eqref{eq:tortoise2}, in these coordinates the waveform of the two equations of motion -- one for the odd parity modes and another for the even -- is given by 
\begin{equation}
\left[-\partial_t^2+\partial_x^2-V_{\mathrm{eff}}(x)\right]\Psi_{\ell m}(x,t) = 0,
\label{eq:wave_tensor_wh}
\end{equation}

where $V_{\mathrm{even}}\equiv[1-r_g/r(|x|)]\mathcal{V}_{\mathrm{even}}$ and $V_{\mathrm{odd}}\equiv[1-r_g/r(|x|)]\mathcal{V}_{\mathrm{odd}}$. Recalling the procedure followed in subsection \ref{subsec:wh}, our setup already includes the effective potentials for the Morris-Thorne wormholes, obtained by reflection of the potential barriers about the ordinate axis. The effective potentials are plotted in the left panel of Fig.~\ref{fig:evenodd_potentials}, which is very similar to the one in the scalar scattering. In the right panel it is possible to notice that the difference between the potentials is only substantial at $(\ell\leq1)$. Both the even and odd solutions of \eqref{eq:tensor_eq_mov} and \eqref{eq:wave_tensor_wh} are already spherical modes used to find the two asymptotic polarizations of the tensor fluctuations propagating in a flat background, such as the term in the diagonal, $h_+$ 
\begin{align}
\displaystyle{h_+=\frac{1}{r(|x|)}\sum_{\ell,m}\bigg\{\Psi_{\ell m}^{\mathrm{even}}\left[\partial^2_{\theta}+\frac{1}{2}\ell(\ell+1)\right]Y_{\ell m}(\theta,\phi)}\nonumber\\
\displaystyle{-\Psi_{\ell m}^{\mathrm{odd}}\frac{i m}{\sin\theta}\left[\partial_{\theta}-\frac{\cos\theta}{\sin\theta}\right]Y_{\ell m}(\theta,\phi)\bigg\}},
\label{eq:hplus}
\end{align} 
and the off-diagonal, $h_{\times}$ 
\begin{align}
\displaystyle{h_{\times}=\frac{1}{r(|x|)}\sum_{\ell,m}\bigg\{\Psi_{\ell m}^{\mathrm{odd}}\left[\partial^2_{\theta}+\frac{1}{2}\ell(\ell+1)\right]Y_{\ell m}(\theta,\phi)}\nonumber\\
\displaystyle{+\Psi_{\ell m}^{\mathrm{even}}\frac{i m}{\sin\theta}\left[\partial_{\theta}-\frac{\cos\theta}{\sin\theta}\right]Y_{\ell m}(\theta,\phi)\bigg\}}.
\label{eq:hx}
\end{align}

From these expressions, it is simple to see that the the monopole ($\ell=0$) and the dipole ($\ell=1$) terms are identically zero. Thus, the first nontrivial contributions come from the quadrupole solutions $\Psi_{20}^{\mathrm{odd}}(x,t)$ and $\Psi_{20}^{\mathrm{even}}(x,t)$, from which the differences in the odd and even potentials are small, and become even smaller for every $\ell>3$, as we can see in right panel of Fig.~\ref{fig:evenodd_potentials}. Additionally, it is reasonable to identify $h_+$ with the even mode and  $h_{\times}$ with the odd in the equatorial plane up to a constant. Hence, our analysis for the scattering dynamics and the reflected/transmitted frequencies does not require from both the even and odd solutions of \eqref{eq:wave_tensor_wh} to extend the discussions from section \ref{sec:scalar}. However, we will explain one of the consequences of the difference between the Regge-Wheeler and Zerilli potentials in Appendix \ref{sec:AppendixPolar}.

@@

\section{Conclusions}\label{sec:conclusions}
In this paper, we studied the scattering of a test scalar and tensor wavepacket on a Morris-Thorne wormhole. Using a Gaussian pulse as an incident initial condition, we showed the time dependent scattering solution of the quadrupole scalar and tensor spherical modes both inside and outside the potential cavities in Figs.~\ref{fig:sol_n_flux_wh} and \ref{fig:even_evolve}, which reflect and transmit throughout the cavity. Furthermore, after finding the transmission and reflection coefficients of the cavities in Fig.~\ref{fig:RT_scalar} and in the left panel of Fig.~\ref{fig:RT_reconst}, we used the geometrical optics approximation to reconstruct the shape of the Fourier transformed asymptotic solutions in Fig.~\ref{fig:rec_scalar} and in the right panel of Fig.~\ref{fig:RT_reconst}. We find that the reconstructed shape of the spectrum is accurate, without showing, however, the QNM peaks.
